Theoretical aspects in gravity modeling

### 2018-04-22 manuscript/letter.tex
finish the anomalous quantities section

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for example) and the normal gravity vector 
at a point $P$. The difference between these two vectors, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
By definition, the centrifugal component of the normal gravity vector 
is equal to the centrifugal component of the gravity vector if they 
are evaluated at the same point.
Then, the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Figure \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the gravity $g_{P}$ and the normal gravity 
$\gamma_{P}$, at the same point $P$, is called \textit{gravity 
disturbance} \citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

@@

PAREI AQUI - precisa descrever as demais quantidade na fig 3

The gravity anomaly is the commonly used quantity in applied 
geophysics. It is defined as the difference
between the gravity on the geoid and the normal gravity on the ellipsoid,
both at the same geodetic latitude and longitude.
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function in space 
\citep{barthelmes2013}.
Different gravity anomalies can be calculated, depending on the
corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
These corrections are usually called gravity reductions.
For example, the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The last one is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the topocentric Cartesian coordinate 
system (Figure \ref{fig:TCS}b).
Consequently, by using the approximation defined in equation 
\ref{eq:gobs-approx}, the gravity disturbance 
(equation \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the topocentric Cartesian coordinate system
(Figure \ref{fig:TCS}).
This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
of the disturbing gravitational attraction $\Delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector. In the topocentric Cartesian coordinate system,
the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})
can be defined as the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source with density contrast $\Delta\rho$
(in $kg / m^{3}$) can be represented by the following harmonic function:
\begin{equation}
d_{i} = k_{g} \, G \, \Delta\rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

>>>>

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for example) and the normal gravity vector 
at a point $P$. 
In this case, the gravity vector $\mitbf{g}_{P}$ represents the 
gradient of a scalar potential called \textit{gravity potential}, 
which is the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the 
gradient of a scalar potential called \textit{normal potential}, 
which is also the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
By definition, the centrifugal part of the normal potential is equal 
to that of the gravity potential.
Then, the both gravity vector and normal gravity vectors are represented 
by the sum of two vectors: the gradient of a gravitational potential and
the gradient of a centrifugal potential. Notice that, as the centrifugal 
part of the gravity potential and normal potential are the same, the
centrifugal part of the gravity vector and normal gravity vector are
also equal to each other.

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
As the centrifugal part of the gravity vector and the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Figure \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between gravity $g_{P}$ (magnitude of the gravity 
vector $\mitbf{g}_{P}$) and normal gravity $\gamma_{P}$ 
(magnitude of the normal gravity vector $\mitbf{\gamma}_{P}$), at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

@@

The gravity anomaly is the commonly used quantity in applied 
geophysics. It is defined as the difference between the 
gravity at a point $Q^{\prime}$ on the geoid and the normal gravity 
at a point $Q$ on the ellipsoid surface,
both at the same geodetic latitude and longitude (Figure \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward of a gravity anomaly.

Different gravity anomalies can be calculated (e.g., Free-air and 
Bouguer anomalies), depending on the corrections applied to 
them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
These corrections are usually called gravity reductions.
It can be shown that the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the topocentric Cartesian coordinate 
system (Figure \ref{fig:TCS}b).
Consequently, by using the approximation defined in equation 
\ref{eq:gobs-approx}, the gravity disturbance 
(equation \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the topocentric Cartesian coordinate system
(Figure \ref{fig:TCS}).
This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
of the disturbing gravitational attraction $\Delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{sanso_sideris2013}. 
In the topocentric Cartesian coordinate system,
the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})
can be defined as the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source with density contrast $\Delta\rho$
(in $kg / m^{3}$) can be represented by the following harmonic function:
\begin{equation}
d_{i} = k_{g} \, G \, \Delta\rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.
